---
title: |
  | Data Exploration and Analysis: 
  | Multiple Regression with Life Expectancy Prediction 
output:
  pdf_document: default
---

```{r message=FALSE, warning=FALSE}

library(readr)
library(leaps)
library(corrplot)

library(tseries)
library(DescTools)
```

## Loading and cleaning dataset

```{r, message=FALSE, warning=FALSE}
# Loading and cleaning data from source

col_names <- c('country', 'year', 'status', 'life_expectancy', 'adult_mortality', 
               'infant_deaths', 'alcohol', 'percent_expend', 'hep_b', 'measles', 
               'bmi', 'deaths_under5', 'polio', 'total_expend', 'diptheria', 'hiv_aids', 
               'gdp', 'population', 'thin_1_19', 'thin_5_9', 'income_comp', 'schooling')

data <- read_csv('data/life_expectancy_raw.csv', 
                 col_names = TRUE, 
                 col_types = cols(population = col_double()), 
                 trim_ws = TRUE)

colnames(data) <- col_names

head(data, 10) #view first 10 rows
```
### Data exploration and scoping

```{r, message=FALSE, warning=FALSE}

# Picking data from one year- try most recent year

data_2015 = data[data$year == 2015, ] # Most recent year available

missing_2015 <- colSums(is.na(data_2015))
print("Columns with more than 10% missing data for 2015:")
print(missing_2015[missing_2015 > 18])

# Two predictors with data mostly incomplete- try the next most recent year

data_2014 = data[data$year == 2014, ] # Next most recent year
missing_2014 <- colSums(is.na(data_2014))
print("Columns with more than 10% missing data for 2014:")
print(missing_2014[missing_2014 > 18])


```
```{r, message=FALSE, warning=FALSE}

# clean out 2014 data with complete columns to use as input for analysis

input <- data_2014[complete.cases(data_2014), ]


# convert country status to numerical dummy variable
input["status_code"] <- NA
input$status_code[input$status == 'Developed'] <- 1
input$status_code[input$status == 'Developing'] <- 0

# drop unused columns for regression model
input <- input[, !(colnames(input) %in% c('country', 'year', 'status'))]

head(input, 10) #view first 10 rows
write.csv(input, file='data/life_expectancy_input.csv')
```

### Check correlation of predictors
```{r, message=FALSE, warning=FALSE}

str(input)
correlation <- cor(input)
corrplot(correlation, method = 'color')

```

Predictors with strong positive correlation with life_expectancy:  

* income_comp
* schooling

Predictors with strong negative correlation with life_expectancy:  

* adult_mortality
* hiv_aids


### Stepwise selection using AIC

```{r, message=FALSE, warning=FALSE}

attach(input)

nullmodel <- lm(life_expectancy~1, data=input)
fullmodel <- lm(life_expectancy~., data=input)

step(nullmodel, data=input, scope=list(upper=fullmodel, lower=nullmodel, direction='both', k=2, test='F'), trace=0)

```
### Reduced linear model after variable selection

After the stepwise selection, we choose to include the following four predictors:  

  * income_comp (income composition of resources index)
  * adult_mortality (adult mortality probability)
  * hiv_aids (death rate from HIV/AIDS)
  * total_expend (government expenditure on health)

From this, we can build the regression model and perform analysis of variance:  

```{r, message=FALSE, warning=FALSE}

model1 <- lm(life_expectancy ~ income_comp + adult_mortality + hiv_aids + total_expend)

anova(model1)
summary(model1)
```

### Testing linear model assumptions

Residual plot for constant variance of residuals:  

```{r,message=FALSE, warning=FALSE}
plot(model1$fit, model1$res, xlab="FItted", ylab="Residuals", main="Residuals Against Fitted Values")

abline(h=0, lty=2)
```

Since there is no observable pattern within the residual plot, the assumption of constant variance is not violated.  

Runs test and Durbin-Watson Test for independence of residuals:  

```{r,message=FALSE, warning=FALSE}

res <- model1$res
runs.test(factor(sign(res)))

DurbinWatsonTest(model1, alternative="two.sided")

```

Since both p-values are sufficiently large, there exists no significant evidence against the null hypothesis of both tests, which suggests that the residuals are not autocorrelated.   

QQ plot for normal distribution of residuals:  

```{r,message=FALSE, warning=FALSE}
qqnorm(res, xlab="normal scores", ylab="ordered residuals")
qqline(res, lty=2)

```

Since the normal probability plot is mostly close to the normal reference line, the assumption that residuals follow a normal distribution is not violated.  